---
title: "Semantic Taxonomic Analysis of Psychological Scales (Wulff & Mata, 2025)"
author: "Ricardo Primi and Leonardo Mose"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Semantic Taxonomic Analysis of Psychological Scales (Wulff & Mata, 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates the semantic taxonomic analyses introduced in
Wulff & Mata (2025), *Escaping the jingle-jangle jungle: increasing conceptual
clarity in psychology using large language models*, as implemented in
**embeddcv**. The workflow asks three questions about a collection of
psychological scales and their construct definitions:

1. **Validation** — do item embeddings recover the intended scale structure
   (*structural fidelity*), do scale embeddings align with their assigned
   definitions (*scale-label alignment*), and how internally consistent is each
   scale semantically (*semantic alpha*)?
2. **Jingle-jangle detection** — which scale pairs have similar labels but
   dissimilar content (*jingle fallacies*), and which have similar content but
   dissimilar labels (*jangle fallacies*)?
3. **Relabeling** — can a data-driven clustering of scale embeddings, combined
   with optimal label assignment, produce a taxonomy with fewer fallacies than
   the original one?

The first part of this vignette runs a small simulated example so all functions
can be executed quickly. The second part documents the full B5/PD pipeline used
with real instrument data.

```{r setup, message = FALSE, warning = FALSE}
library(embeddcv)
library(dplyr)
library(ggplot2)

set.seed(42)
```

## Part 1: A small simulated example

We simulate embeddings for items belonging to five Big Five constructs, each
measured by two instruments ("bfi" and "ipip"). Items from the same construct
share a common center in embedding space.

```{r simulate}
n_dim <- 50
constructs <- c("openness", "conscientiousness", "extraversion",
                "agreeableness", "neuroticism")
centers <- matrix(rnorm(5 * n_dim), 5, n_dim)

make_emb <- function(n, center) center + matrix(rnorm(n * n_dim, sd = .35), n, n_dim)

# 8 items per construct (4 per instrument)
item_emb <- do.call(rbind, lapply(1:5, function(i) make_emb(8, centers[i, ])))
rownames(item_emb) <- paste0("it_", rep(constructs, each = 8), "_", 1:8)
item_scale_id <- paste0(rep(c("bfi", "ipip"), 20), "_", rep(constructs, each = 8))

# Definition/label embeddings: one definition per scale
label_emb <- do.call(rbind, lapply(1:5, function(i) make_emb(2, centers[i, ])))
rownames(label_emb) <- paste0(rep(c("bfi", "ipip"), times = 5),
                              "_", rep(constructs, each = 2))
```

### Scale embeddings and similarity matrices

`get_scale_embeddings()` averages item embeddings into one vector per scale.
`cosine_matrix()` computes within-set similarities, and `rescale_similarity()`
applies the min-max rescaling used by Wulff & Mata before clustering.

```{r matrices}
scale_emb <- get_scale_embeddings(item_emb, item_scale_id)

scale_cos_raw <- cosine_matrix(scale_emb)
scale_cos     <- rescale_similarity(scale_cos_raw)
label_cos     <- cosine_matrix(label_emb)

# Original taxonomy: each scale keeps its own definition
label_by_scale <- setNames(rownames(scale_emb), rownames(scale_emb))
label_by_scale <- label_by_scale[rownames(scale_cos)]

round(scale_cos_raw[1:4, 1:4], 2)
```

### All pairwise comparisons at once

`cosim_all_pairs()` wraps the generalized `cosim_itens_scales()` to produce
item-item, item-scale, item-definition, scale-scale, scale-definition, and
definition-definition similarities in one call.

```{r all-pairs}
sims <- cosim_all_pairs(
  embeddings = list(item = item_emb, scale = scale_emb, definition = label_emb),
  make_plot = FALSE
)
names(sims)
head(sims$scale_definition$cosim_long)
```

### Validation: semantic alpha and structural fidelity

```{r validation}
item_cos <- cosine_matrix(item_emb)

semantic_alpha_tbl <- semantic_alpha(
  item_cos = item_cos,
  scale_id = item_scale_id,
  model_names = "toy"
)
semantic_alpha_tbl

structural_fidelity_tbl <- structural_fidelity(
  item_cos = item_cos,
  scale_id = item_scale_id,
  inventory = sub("_.*", "", item_scale_id),
  model_names = "toy"
)
structural_fidelity_tbl
```

```{r fidelity-plot, fig.height = 4}
plot_structural_fidelity(structural_fidelity_tbl, show_legend = FALSE)
```

### Jingle-jangle detection

`jingle_jangle_pairs()` builds a pair-level table combining scale and label
similarities; `jingle_jangle_thresholds()` computes the Wulff & Mata cutoff
rule (low = 1 - 1/5, high = 1 - 1/100); `classify_jingle_jangle()` flags each
pair.

```{r jingle-jangle}
jj_base <- jingle_jangle_pairs(
  scale_cos = scale_cos,
  label_cos = label_cos,
  label_by_scale = label_by_scale
)

thresholds <- jingle_jangle_thresholds(scale_cos = scale_cos, label_cos = label_cos)
thresholds

jj <- classify_jingle_jangle(jj_base, thresholds, verbose = FALSE)
count(jj, type)
```

```{r jj-scatter, fig.height = 5.5}
plot_jingle_jangle_scatter(pair_table = jj, thresholds = thresholds)
```

### Relabeling: clustering + optimal label assignment

`make_relabeling_clustering_solutions()` generates candidate taxonomies
(hierarchical, k-means, and optionally Mclust/Louvain).
`score_relabeling_solutions()` assigns the best label to each cluster via
bipartite matching on the scale-definition similarity matrix and counts the
fallacies of each candidate. `select_relabeling_solutions()` applies the
parsimonious/optimal selection rule.

```{r relabeling}
scale_label_cross <- cosim_itens_scales(
  item_emb = scale_emb,
  scale_emb = label_emb,
  x_type = "scale",
  y_type = "definition",
  make_plot = FALSE
)$cosim_matrix

solutions <- make_relabeling_clustering_solutions(
  scale_cos = scale_cos,
  k_seq = 2:6,
  include_mclust = FALSE,
  include_louvain = FALSE,
  seed = 1
)

scored <- score_relabeling_solutions(
  clustering_solutions = solutions,
  scale_label_cross = scale_label_cross,
  scale_cos = scale_cos,
  label_cos = label_cos,
  thresholds = thresholds
)

select_relabeling_solutions(labels_scored = scored, max_constructs = 10)
```

The original taxonomy can be scored the same way, providing the baseline that
relabeling tries to improve:

```{r original-score}
score_label_solution(
  solution = label_by_scale,
  scale_cos = scale_cos,
  label_cos = label_cos,
  thresholds = thresholds
)
```

## Part 2: The full B5/PD pipeline

The complete analysis of Big Five and personality-disorder instruments uses
saved item and definition embeddings (`dados.RData`, ~33 MB, not shipped with
the package; contact the authors). The code below condenses the full pipeline
script `b5_pd_wulff_mata_embeddcv.R`; it is shown but not evaluated.

```{r full-pipeline, eval = FALSE}
load("dados.RData")   # itens, definicoes, item_embedd, definicoes_embedd

# 1. Prepare matrices -------------------------------------------------------
item_emb <- as.matrix(item_embedd)
rownames(item_emb) <- make.unique(as.character(itens$cod_item))

# Harmonize scale ids between the item table and the definition table,
# then group item embeddings into scale embeddings.
item_scale_id <- itens$scale   # after the id-harmonization steps
scale_emb <- get_scale_embeddings(item_emb, item_scale_id)

label_emb <- as.matrix(definicoes_embedd)
rownames(label_emb) <- definicoes$scale
label_by_scale <- setNames(definicoes$scale, definicoes$scale)

# 2. Similarity matrices ----------------------------------------------------
scale_cos_raw <- cosine_matrix(scale_emb)
scale_cos     <- rescale_similarity(scale_cos_raw)
label_cos     <- cosine_matrix(label_emb)

scale_label_cross <- cosim_itens_scales(
  item_emb = scale_emb, scale_emb = label_emb,
  x_type = "scale", y_type = "definition", make_plot = FALSE
)$cosim_matrix

# Diagnostic distributions for every embedding-level comparison
sims <- cosim_all_pairs(
  embeddings = list(item = item_emb, scale = scale_emb, definition = label_emb),
  make_plot = FALSE
)
plot_similarity_distribution(
  similarities = sims$scale_scale$cosim_long,
  title = "Scale-scale cosine similarity distribution",
  show_quartiles = TRUE
)

# 3. Validation -------------------------------------------------------------
item_cos <- cosine_matrix(item_emb)

semantic_alpha_tbl <- semantic_alpha(item_cos, item_scale_id, model_names = "B5")

structural_fidelity_tbl <- structural_fidelity(
  item_cos = item_cos,
  scale_id = item_scale_id,
  inventory = itens$instrumento,
  model_names = "B5"
)
plot_structural_fidelity(structural_fidelity_tbl,
                         output_file = "01A_structural_fidelity.png")

scale_label_alignment_tbl <- scale_label_alignment(
  scale_emb = scale_emb,
  label_embeddings = list(B5 = label_emb),
  label_by_scale = label_by_scale[rownames(scale_emb)]
)
plot_scale_label_alignment(scale_label_alignment_tbl)

convergent_validity_tbl <- convergent_validity(
  scale_cos = scale_cos_raw,
  construct_by_scale = scale_label_from_scale(label_by_scale[rownames(scale_cos_raw)]),
  min_scales_per_construct = 3
)
plot_convergent_validity(convergent_validity_tbl)

# 4. Jingle-jangle ----------------------------------------------------------
jj_base    <- jingle_jangle_pairs(scale_cos, label_cos, label_by_scale)
thresholds <- jingle_jangle_thresholds(scale_cos, label_cos,
                                       min_constructs = 5, max_constructs = 100)
jj         <- classify_jingle_jangle(jj_base, thresholds)

plot_jingle_jangle_scatter(jj, thresholds, output_file = "02A_scatter.png")
plot_jingle_jangle(jj, thresholds, output_file = "03_jingle_jangle.png")

# Sensitivity of fallacy counts to the cutoff rule
grid <- jingle_jangle_threshold_grid(jj_base, scale_cos, label_cos,
                                     lows = 1:10, highs = seq(50, 250, 20))
plot_jingle_jangle_cutoffs(grid, output_file = "S01_cutoffs.png")

# 5. Relabeling -------------------------------------------------------------
original_solution <- label_by_scale[rownames(scale_cos)]
original_score <- score_label_solution(original_solution, scale_cos, label_cos,
                                       thresholds)

solutions <- make_relabeling_clustering_solutions(
  scale_cos = scale_cos,
  k_seq = seq_len(min(nrow(scale_cos), nrow(label_cos))),
  include_mclust = TRUE,
  include_louvain = TRUE,
  seed = 42
)

scored <- score_relabeling_solutions(solutions, scale_label_cross,
                                     scale_cos, label_cos, thresholds)

selection <- select_relabeling_solutions(scored, max_constructs = 100)

# 6. Figures ----------------------------------------------------------------
panels <- prepare_relabeling_panels(
  labels_scored = scored,
  scale_label_cross = scale_label_cross,
  label_by_scale = label_by_scale,
  max_constructs = 100,
  palette = "magma"
)
plot_relabeling_figure(panels, output_file = "04_relabeling.png",
                       max_constructs = 100, palette = "magma")
plot_relabeling_robustness(scored,
                           original_fallacy_sum = original_score["fallacy_sum"],
                           output_file = "S02_robustness.png")
```

## References

- Wulff, D. U., & Mata, R. (2025). Escaping the jingle-jangle jungle:
  increasing conceptual clarity in psychology using large language models.
  *Nature Human Behaviour*. doi:10.1038/s41562-024-02089-y
- Hommel, B. E., & Arslan, R. C. (2024). Language models accurately infer
  correlations between psychological items and scales from text alone.
  doi:10.31234/osf.io/kjuce
