---
title: "Semantic Taxonomic Analysis of Psychological Scales (Wulff & Mata, 2025)"
author: "Ricardo Primi and Leonardo Mose"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Semantic Taxonomic Analysis of Psychological Scales (Wulff & Mata, 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates the semantic taxonomic analyses introduced in
Wulff & Mata (2025), *Escaping the jingle-jangle jungle: increasing conceptual
clarity in psychology using large language models*, as implemented in
**embeddcv**. The workflow asks three questions about a collection of
psychological scales and their construct definitions:

1. **Validation** — `semantic_validity()` summarises, per scale, a semantic
   reliability (*alpha*), *structural fidelity* (items closer within their
   scale than to other scales), and, when constructs are supplied, *convergent
   validity* (scales of the same construct closer to one another).
2. **Jingle-jangle detection** — which scale pairs have similar labels but
   dissimilar content (*jingle fallacies*), and which have similar content but
   dissimilar labels (*jangle fallacies*)?
3. **Relabeling** — can a data-driven clustering of scale embeddings, combined
   with optimal label assignment, produce a taxonomy with fewer fallacies than
   the original one?

The functions operate on **one instrument** at a time; to compare instruments
you run the analysis per instrument in your own pipeline. The first part of
this vignette runs a small simulated example so all functions can be executed
quickly; the second part documents the full B5/PD pipeline.

```{r setup, message = FALSE, warning = FALSE}
library(embeddcv)
library(dplyr)
library(ggplot2)

set.seed(42)
```

## Part 1: A small simulated example

We simulate embeddings for items belonging to five Big Five constructs. Items
from the same construct share a common center in embedding space. Each construct
is measured by two scales, so convergent validity can be illustrated.

```{r simulate}
n_dim <- 50
constructs <- c("openness", "conscientiousness", "extraversion",
                "agreeableness", "neuroticism")
centers <- matrix(rnorm(5 * n_dim), 5, n_dim)

make_emb <- function(n, center) center + matrix(rnorm(n * n_dim, sd = .35), n, n_dim)

# 8 items per construct (4 per scale, two scales per construct)
item_emb <- do.call(rbind, lapply(1:5, function(i) make_emb(8, centers[i, ])))
rownames(item_emb) <- paste0("it_", rep(constructs, each = 8), "_", 1:8)
item_scale_id <- paste0(rep(c("s1", "s2"), 20), "_", rep(constructs, each = 8))

# One definition/label embedding per scale
label_emb <- do.call(rbind, lapply(1:5, function(i) make_emb(2, centers[i, ])))
rownames(label_emb) <- paste0(rep(c("s1", "s2"), times = 5),
                              "_", rep(constructs, each = 2))
```

### Scale embeddings and similarity matrices

`get_scale_embeddings()` averages item embeddings into one vector per scale.
`cosine_matrix()` computes within-set similarities, and `rescale_similarity()`
applies the min-max rescaling used by Wulff & Mata before clustering.

```{r matrices}
scale_emb <- get_scale_embeddings(item_emb, item_scale_id)

scale_cos_raw <- cosine_matrix(scale_emb)
scale_cos     <- rescale_similarity(scale_cos_raw)
label_cos     <- cosine_matrix(label_emb)

# Original taxonomy: each scale keeps its own definition
label_by_scale <- setNames(rownames(scale_emb), rownames(scale_emb))
label_by_scale <- label_by_scale[rownames(scale_cos)]

round(scale_cos_raw[1:4, 1:4], 2)
```

### Cosine similarities between any two sets

`cosim_xy()` is the general engine: pass two embedding sets for a cross
comparison, or one set (`y = NULL`) for a symmetric within-set matrix. It
returns the raw `$matrix`, a tidy `$long` pair table, and an optional `$plot`.
`cosim_all_pairs()` wraps it to produce every pairwise comparison at once.

```{r all-pairs}
sims <- cosim_all_pairs(
  embeddings = list(item = item_emb, scale = scale_emb, definition = label_emb),
  make_plot = FALSE
)
names(sims)
head(sims$scale_definition$long)
```

### Validation: one tidy table per scale

`semantic_validity()` computes the semantic alpha, structural fidelity, and
(given a construct grouping) convergent validity for every scale in one call.

```{r validation}
# Map each scale to its construct (drop the "s1_"/"s2_" prefix)
construct_by_scale <- setNames(
  sub("^s[12]_", "", rownames(scale_emb)),
  rownames(scale_emb)
)

validity <- semantic_validity(
  item_emb = item_emb,
  scale_id = item_scale_id,
  construct_by_scale = construct_by_scale,
  min_scales = 3
)
validity
```

```{r fidelity-plot, fig.height = 4}
plot_semantic_validity(validity, index = "fidelity")
```

### Jingle-jangle detection

`jingle_jangle_pairs()` builds a pair-level table combining scale and label
similarities; `jingle_jangle_thresholds()` computes the Wulff & Mata cutoff
rule (low = 1 - 1/5, high = 1 - 1/100); `classify_jingle_jangle()` flags each
pair.

```{r jingle-jangle}
jj_base <- jingle_jangle_pairs(
  scale_cos = scale_cos,
  label_cos = label_cos,
  label_by_scale = label_by_scale
)

thresholds <- jingle_jangle_thresholds(scale_cos = scale_cos, label_cos = label_cos)
thresholds

jj <- classify_jingle_jangle(jj_base, thresholds, verbose = FALSE)
count(jj, type)
```

```{r jj-scatter, fig.height = 5.5}
plot_jingle_jangle_scatter(pair_table = jj, thresholds = thresholds)
```

### Relabeling: clustering + optimal label assignment

`make_relabeling_clustering_solutions()` generates candidate taxonomies.
`score_relabeling_solutions()` assigns the best label to each cluster via
bipartite matching on the scale-definition similarity matrix (from
`cosim_xy(...)$matrix`) and counts the fallacies of each candidate.
`select_relabeling_solutions()` applies the parsimonious/optimal selection rule.

```{r relabeling}
scale_label_cross <- cosim_xy(
  scale_emb, label_emb,
  x_type = "scale", y_type = "definition"
)$matrix

solutions <- make_relabeling_clustering_solutions(
  scale_cos = scale_cos,
  k_seq = 2:6,
  include_mclust = FALSE,
  include_louvain = FALSE,
  seed = 1
)

scored <- score_relabeling_solutions(
  clustering_solutions = solutions,
  scale_label_cross = scale_label_cross,
  scale_cos = scale_cos,
  label_cos = label_cos,
  thresholds = thresholds
)

select_relabeling_solutions(labels_scored = scored, max_constructs = 10)
```

The original taxonomy can be scored the same way, providing the baseline that
relabeling tries to improve:

```{r original-score}
score_label_solution(
  solution = label_by_scale,
  scale_cos = scale_cos,
  label_cos = label_cos,
  thresholds = thresholds
)
```

## Part 2: The full B5/PD pipeline

The complete analysis of Big Five and personality-disorder instruments uses
saved item and definition embeddings (`dados.RData`, ~33 MB, not shipped with
the package; contact the authors). The code below condenses the full pipeline;
it is shown but not evaluated. For several instruments, run the validation per
instrument and combine the resulting tables in your own pipeline.

```{r full-pipeline, eval = FALSE}
load("dados.RData")   # itens, definicoes, item_embedd, definicoes_embedd

# 1. Prepare matrices -------------------------------------------------------
item_emb <- as.matrix(item_embedd)
rownames(item_emb) <- make.unique(as.character(itens$cod_item))

item_scale_id <- itens$scale   # after id-harmonization
scale_emb <- get_scale_embeddings(item_emb, item_scale_id)

label_emb <- as.matrix(definicoes_embedd)
rownames(label_emb) <- definicoes$scale
label_by_scale <- setNames(definicoes$scale, definicoes$scale)

# 2. Similarity matrices ----------------------------------------------------
scale_cos_raw <- cosine_matrix(scale_emb)
scale_cos     <- rescale_similarity(scale_cos_raw)
label_cos     <- cosine_matrix(label_emb)

# scale x definition cross-similarity (weights for relabeling)
scale_label_cross <- cosim_xy(
  scale_emb, label_emb,
  x_type = "scale", y_type = "definition"
)$matrix

# Diagnostic distributions for every embedding-level comparison
sims <- cosim_all_pairs(
  embeddings = list(item = item_emb, scale = scale_emb, definition = label_emb),
  make_plot = FALSE
)
plot_similarity_distribution(
  similarities = sims$scale_scale$long,
  title = "Scale-scale cosine similarity distribution",
  show_quartiles = TRUE
)

# 3. Validation (one instrument) --------------------------------------------
# construct_by_scale maps each scale to its construct (named by scale id).
validity <- semantic_validity(
  item_emb = item_emb,
  scale_id = item_scale_id,
  construct_by_scale = setNames(definicoes$construct, definicoes$scale)
)
plot_semantic_validity(validity, index = "fidelity",
                       output_file = "01A_fidelity.png")
plot_semantic_validity(validity, index = "alpha_semantic", recovery_line = NULL)
plot_semantic_validity(validity, index = "convergent_similarity")

# Divergent validity needs an EXTERNAL criterion (observed correlations):
# divergent_validity(similarity = scale_cos_raw, empirical_pairs = obs_pairs,
#                    id_1_col = "scale_1", id_2_col = "scale_2",
#                    observed_col = "abs_cor")

# 4. Jingle-jangle ----------------------------------------------------------
jj_base    <- jingle_jangle_pairs(scale_cos, label_cos, label_by_scale)
thresholds <- jingle_jangle_thresholds(scale_cos, label_cos,
                                       min_constructs = 5, max_constructs = 100)
jj         <- classify_jingle_jangle(jj_base, thresholds)

plot_jingle_jangle_scatter(jj, thresholds, output_file = "02A_scatter.png")
plot_jingle_jangle(jj, thresholds, output_file = "03_jingle_jangle.png")

grid <- jingle_jangle_threshold_grid(jj_base, scale_cos, label_cos,
                                     lows = 1:10, highs = seq(50, 250, 20))
plot_jingle_jangle_cutoffs(grid, output_file = "S01_cutoffs.png")

# 5. Relabeling -------------------------------------------------------------
original_solution <- label_by_scale[rownames(scale_cos)]
original_score <- score_label_solution(original_solution, scale_cos, label_cos,
                                       thresholds)

solutions <- make_relabeling_clustering_solutions(
  scale_cos = scale_cos,
  k_seq = seq_len(min(nrow(scale_cos), nrow(label_cos))),
  include_mclust = TRUE,
  include_louvain = TRUE,
  seed = 42
)

scored <- score_relabeling_solutions(solutions, scale_label_cross,
                                     scale_cos, label_cos, thresholds)

selection <- select_relabeling_solutions(scored, max_constructs = 100)

# 6. Figures ----------------------------------------------------------------
panels <- prepare_relabeling_panels(
  labels_scored = scored,
  scale_label_cross = scale_label_cross,
  label_by_scale = label_by_scale,
  max_constructs = 100,
  palette = "magma"
)
plot_relabeling_figure(relabeling_panels = panels, output_file = "04_relabeling.png",
                       max_constructs = 100, palette = "magma")
plot_relabeling_robustness(scored,
                           original_fallacy_sum = original_score["fallacy_sum"],
                           output_file = "S02_robustness.png")
```

## References

- Wulff, D. U., & Mata, R. (2025). Escaping the jingle-jangle jungle:
  increasing conceptual clarity in psychology using large language models.
  *Nature Human Behaviour*. doi:10.1038/s41562-024-02089-y
- Hommel, B. E., & Arslan, R. C. (2024). Language models accurately infer
  correlations between psychological items and scales from text alone.
  doi:10.31234/osf.io/kjuce
